---
title: 'Explore income inequality in Canadian cities: an interactive map of 2001-
  2016 Gini coefficient data'
author: "Megan Wylie"
date: '2020-01-02'
output: pdf_document
tags:
- R Markdown
- plot
categories: R
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F, message = F, warning = F)
# echo = FALSE doesn't show the R code in the post.
# eval = FALSE to omit results

library(tidyverse)
library(geojsonio)
library(sf)
#library(mapview)
library(leaflet)
library(shiny)
library(rgdal)

```

# Background: income inequality and health research

I completed master's research in April of 2019 on the topic of income inequality and its impacts on health in Canadian cities. As part of this project, I calculated income inequality in 130 cities for 2001 through 2011. I was also able to calculate income inequality in 2016 once the census data was released in 2018. 

I've now repurposed these data into an interactive leaflet map for your exploration! 

# How to use the map

The below interactive map shows cities in Canada. The size of each citcle represents the level of income inequality. 

Income inequality here is measured by the Gini coefficient, a measure that compares the incomes of everyone who lives in that city to a theoretical, perfectly equal, city. In this sense, the Gini coefficient measures how far each city is from perfect equality. The closer to 1, the more unequal the city.

Cities in Canada generally range from low 0.3 score to higher 0.5. As a comparison to other countries, Boston had a Gini coefficient measure of .48 in 2015 and Oslo .31.

I calculated the income inequality bins by percentile. For each year of analysis, I binned the cities by the lowest 25%, highest 25%, the 25-50th percentile, and the 50-75th percentile. 

*You can zoom in and out on the map, and select different layers of data from the top left layer options. When dragging your mouse over a city, the city name and Gini coefficient will appear.* 


```{r map, echo = F, message = FALSE}
# Name cities
# mapview::mapview(point.locations.sf, label = point.locations.sf$CMANAME)

point.locations.sf <- read_rds("canada_city_inequality_sf.RDS")

point.locations.sf.transform <- sf::st_transform(point.locations.sf , 4326)

# make quintiles to bin the gini coefficients
tf <- point.locations.sf.transform %>% 
  left_join(point.locations.sf.transform %>% 
              as.tibble() %>% 
              select(CMANAME , gini01 , gini06 , gini11 , gini16) %>% 
              mutate_at(c("gini01" , "gini06" , "gini11" , "gini16") , 
                        dplyr::percent_rank) %>% 
              mutate(gini01_bin = if_else(gini01 > .75 , 4 , 
                                          if_else(gini01 > .5 , 3 , 
                                                  if_else(gini01 > .25 , 2 , 1))), 
                     gini06_bin = if_else(gini06 > .75 , 4 , 
                                          if_else(gini06 > .5 , 3 , 
                                                  if_else(gini06 > .25 , 2 , 1))),
                     gini11_bin = if_else(gini11 > .75 , 4 , 
                                          if_else(gini11 > .5 , 3 , 
                                                  if_else(gini11 > .25 , 2 , 1))),
                     gini16_bin = if_else(gini16 > .75 , 4 , 
                                          if_else(gini16 > .5 , 3 , 
                                                  if_else(gini16 > .25 , 2 , 1))),
              ) %>% select(CMANAME , gini01_bin , gini06_bin
                           , gini11_bin , gini16_bin) , 
            by = "CMANAME")

#set up legend - css/html styles
# source: https://stackoverflow.com/questions/58505589/circles-in-legend-for-leaflet-map-with-addcirclemarkers-in-r-without-shiny

addLegendCustom <- function(map, colors, title, labels, sizes, opacity = 0.6, position = "topright"){
  colorAdditions <- paste0(colors, "; border-radius: 50%; width:", 
                           sizes, "px; height:", sizes, "px")
  labelAdditions <- paste0("<div style='display: inline-block;text-align: right;height: ", 
                           sizes, "px;margin-top: 4px;line-height: ", sizes, "px;'>", 
                           labels, "</div>")
  
  return(
    addLegend(map,
      colors = colorAdditions, 
      labels = labelAdditions,
      opacity = opacity,
      title = title,
      position = position
    )
  )
}

# map via leaflet

leaflet(tf) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  #addCircleMarkers defines size in terms of metres - they're too similar in size
  addCircleMarkers(color = "#3182bd" ,
                   fillColor = "#9ecae1" ,
                   fillOpacity = .6 , 
                   radius = ~ gini01_bin*2 ,
                   weight = 2 ,
                   label = ~paste(CMANAME , round(gini01, digits = 2) , sep = ", " ),
                   group = "Gini Coefficient 2001") %>% 
  addCircleMarkers( color = "#3182bd" ,
                    fillColor = "#9ecae1" ,
                    fillOpacity = .6 , 
                    radius = ~ gini06_bin*2 ,
                    weight = 2 ,
                    label = ~paste(CMANAME , round(gini06, digits = 2) , sep = ", "),
                    group = "Gini Coefficient 2006") %>% 
  addCircleMarkers( color = "#3182bd" ,
                  fillColor = "#9ecae1" ,
                  fillOpacity = .6 , 
                  radius = ~ gini11_bin*2 ,
                  weight = 2 ,
                  label = ~paste(CMANAME , round(gini11, digits = 2) , sep = ", " ) ,
                  group = "Gini Coefficient 2011") %>%
  addCircleMarkers( color = "#3182bd" ,
                    fillColor = "#9ecae1" ,
                    fillOpacity = .6 , 
                    radius = ~ gini16_bin*2 ,
                    weight = 2 ,
                    label = ~paste(CMANAME , round(gini16, digits = 2), sep = ", " ),
                    group = "Gini Coefficient 2016") %>%
  #addLayersControl command breaks each year into one selection option.
  addLayersControl(
                overlayGroups = c("Gini Coefficient 2001", "Gini Coefficient 2006" ,
                             "Gini Coefficient 2011" , "Gini Coefficient 2016")) %>% 
  #add legends that appear/disappear with toggle
  addLegendCustom(colors = ("#3182bd"),
    labels = c("< 25th percentile",
               "50th percentile", 
               "75th percentile", 
               "> 75th percentile"),
    sizes = c(5, 10, 15, 20),
    title = "Urban Income Inequality Level") %>%
  #hide layers 2001, 2006, and 2011 for when the map opens. 
  hideGroup("Gini Coefficient 2001") %>% 
              hideGroup("Gini Coefficient 2006") %>% 
            hideGroup("Gini Coefficient 2011")

```


# More information

Thank you for your interest in my work! For more information on the data and calculations of the Gini coefficients, see my [thesis publication](http://digitool.library.mcgill.ca/R/EENC3B6JRGRNQYTFUC1QL1K64F9UY9YDIC1UPKXTDG24PQG39G-07956).

In the following months I will be working to publish the research in an academic journal, which I will also link here. 


Megan 
