---
title: Analyzing stock value with R Ladies, Part 1
author: Megan Wylie
date: '2020-01-21'
slug: analyzing-stock-value-with-r-ladies
categories: []
tags: ["stock market", "R", "PortfolioAnalytics", "portfolio" ]
Categories: []
Description: ''
Tags: []
draft: TRUE
---

This week for R Ladies we did a tutorial on stock market analysis. 

This first R chunk sets up the R-Markdown environment. Within the brackets of the first line, the R code is called *setup*. Include = FALSE (or F) means that when the R-Markdown file is run from R Studio to another file format, i.e. being *knit*, it does not run the code.

This tutorial is longer and more complex than the previous workshop, and I've split it into two blog posts. 

In the first blog post, I review:
- getting stock data from Yahoo
- x
- y

In the second blog post, I show you how to:
- z
- a
- b

This tutorial also requires a few new packages, loaded below. (Notice DPLYR from the last workshop!)

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F, message = F, warning = F)

library(PortfolioAnalytics)
library(PerformanceAnalytics)
library(quantmod)
library(zoo)
library(plotly)
library(tidyquant)
library(dplyr)
library(DEoptim)
library(ROI)
require(ROI.plugin.glpk)
require(ROI.plugin.quadprog)
library(xts)

```

For this tutorial, I am going to explain the code by chunk, which I have separated by <- thing.

```{r , echo = F, message = FALSE}
# this gets data from the stock exchange.
getSymbolData <- function(symbolList, start = "2019-06-01", end = "2019-12-31"){
  getSymbols(
    symbolList, 
    auto.assign = TRUE,
    from = start, 
    to = end,
  )
  # Assign to dataframe
  # Get adjusted prices
  p_data <- xts(
    get(symbolList[1])[, 6]
  )
  for (i in 2:length(symbolList)){
    p_data <- merge.xts(p_data, get(symbolList[i])[, 6])
  }
  colnames(p_data) <- symbolList
  # calculate returns
  r_data <- na.omit(CalculateReturns(p_data))
  colnames(r_data) <- symbolList
  list(returns_data = r_data, price_data = p_data)
}
```

We chose to get data from BNS, Air Canada, Shopify, and GoEasy. These were arbitrary choices, and you can find the companies stock code by just googling *stock [company name]*

```{r , echo = F, message = FALSE}
# Stocks BNS Air Canada, Shopify, goeasy
my_stocks <- getSymbolData(c("BNS.TO", "AC", "SHOP", "GSY"))
# save(x = my_stocks, file = "my_stocks.RDS")
``` 




```{r , echo = F, message = FALSE}
#Megan - this makes it a data frame that I can look at of return data.  
return_data <- as.data.frame(my_stocks$returns_data)
View(return_data)
``` 



```{r , echo = F, message = FALSE}
# megan - this changes the name to BNS_TO
names(return_data) <- c("BNS_TO", "AC", "SHOP", "GSY")
View(return_data)

#megan - I think that it's saving different parts of the data in different data frames. 
``` 




```{r , echo = F, message = FALSE}
#return_data$Date <- index(my_stocks$returns_data)
price_data <- as.data.frame(my_stocks$price_data)
#View(price_data)
names(price_data) <- c("BNS_TO", "AC", "SHOP", "GSY")
``` 



```{r , echo = F, message = FALSE}
# this makes a "value"? XTS data.
DATE = index(my_stocks$returns_data)
#"structure"
#str(my_stocks)
#View(my_stocks)

my_stocks$returns_data
``` 




```{r , echo = F, message = FALSE}
# Look at plot
plotly::plot_ly(
  data = price_data,
  y = ~ BNS_TO,
  type = "scatter",
  mode = "lines"
) %>% layout(
  title = title,
  xaxis = list(
    type = "date"
  ),
  yaxis = list(
    title = "Price"
  )
)
``` 


# NEW POST - Post #2 
- looking at the portfolio.

```{r , echo = F, message = FALSE}
returns <- my_stocks$returns_data
price <- my_stocks$price_data
``` 



```{r , echo = F, message = FALSE}
# Mean returns and covariance matrix
mean_returns <- colMeans(returns)
cov_matrix <- cov(returns)

# side note - ?? and ? before search for what a function means.
``` 



```{r , echo = F, message = FALSE}
# Portfolio specification
my_folio <- portfolio.spec(assets = colnames(returns))
``` 



```{r , echo = F, message = FALSE}
# Add constraint 
my_folio <- add.constraint(
  portfolio = my_folio, 
  type = "box", 
  min = 0.05, #these are risk
  max = 0.65
)
``` 





```{r , echo = F, message = FALSE}
my_folio <- add.constraint(
  portfolio = my_folio, 
  type = "leverage", # "full_investment"
  min_sum = 0.99,  #this is very risky
  max_sum = 1.01
) 
``` 


# post 3 - maximizing return! 

```{r , echo = F, message = FALSE}
# Maximize the mean return with ROI
max_return <- add.objective(
  portfolio = my_folio, 
  type = "return", #return
  name = "mean"
) 
``` 



```{r , echo = F, message = FALSE}
# Optimization
opt_max_return <- optimize.portfolio(
  R = returns, 
  portfolio = max_return, 
  optimize_method = "ROI",
  trace = TRUE
)
``` 



```{r , echo = F, message = FALSE}
# these print the details on how much I should invest.
print.default(opt_max_return) #this prints a lot more than the other one.. (after)
print(opt_max_return)
``` 



```{r , echo = F, message = FALSE}
# Plot optimized maximum mean (i.e. return?)
plot(
  opt_max_return, 
  risk.col = "StdDev", 
  return.col = "mean",
  main = "Minimum Variance Optimization", 
  chart.assets = TRUE,
  xlim = c(-0.5, 1), 
  ylim = c(-0.002, 0.004)
)
``` 



```{r , echo = F, message = FALSE}
# Minimize variance (risk) with ROI
min_var <- add.objective(
  portfolio = my_folio, 
  type = "risk", #risk
  name = "var"
)
``` 



```{r , echo = F, message = FALSE}
# Optimize. Note that although ’var’ is the risk metric, 
# ’StdDev’ is returned as an objective measure.
opt_min_var <- optimize.portfolio(
  R = returns, 
  portfolio = min_var,
  optimize_method = "ROI",  
  trace = TRUE
)
``` 




```{r , echo = F, message = FALSE}
print.default(opt_min_var)
print(opt_min_var)
``` 



```{r , echo = F, message = FALSE}
# Plot optimized minimum var (i.e. risk)
plot(
  opt_min_var, 
  risk.col = "StdDev", 
  return.col = "mean",
  main = "Minimum Variance Optimization", 
  chart.assets = TRUE,
  xlim = c(-0.5, 1), 
  ylim = c(-0.002, 0.004)
)
``` 

# post 4 - efficient frontier! 

```{r , echo = F, message = FALSE}
meansd_ef <- create.EfficientFrontier(
  R = returns,
  portfolio = my_folio,
  type = "mean-sd",
  n.portfolios = 10,
)
``` 



```{r , echo = F, message = FALSE}
# Plot efficient frontier
chart.EfficientFrontier(
  meansd_ef,
  match.col = "StdDev", # which column to use for risk
  type = "l", 
  RAR.text = "Sharpe Ratio",
  tangent.line = FALSE,
  chart.assets = TRUE,
  labels.assets = TRUE,
  xlim = c(-0.5, 1), 
  ylim = c(-0.002, 0.004),
  element.color = "blue"
)

``` 

